Algorithmic Approaches to Match Degraded Land Impressions

### 2017-10-03 response.Rmd

---
title: "Response to reviewers"
author: "Eric Hare, Heike Hofmann, Alicia Carriquiry"
date: "October 5th, 2017"
output:
  pdf_document: default
bibliography: references.bib
---

# Reviewer 1

1. *pp. 4  first paragraph of section “3 Model Training”, the author state: "Thus, every Cary bullet is a known match to every other Cary bullet, although early firings do not produce the same markings that later firings do." The authors did not study this here or elsewhere. The authors also did not cite a study that indicated this conclusion. The second half of that statement may very well be true, but  stating it anecdotally should not appear in a research paper.  I’d suggest the minor change: "Thus every Cary bullet is a known match to every other Cary bullet"*

  **Thank you for the comment. We have made this change.**

2. *pp8-9, 11, 25: The authors highlight operator effect on the data acquisition process in the study. The instrumentation used to acquire data requires skill and experience. Naturally that brings up the question as to who was operating the microscope? The authors must state (in at least rough terms) the level of expertise of the operator(s). Were they students new to the technology? Were the operators familiar with the quirks of the surfaces they were looking at? Were they professional surface metrologists or company technicians? I would expect significant operator effects if data is collected by operators inexperienced  with the instrumentation, surface metrology and/or firearms analysis.*

  **In this case, the operators were experienced surface metrologists at NIST. The presence of operator effects was not a conclusion meant to insinuate that these experts performed the scans poorly, but rather that there is variability in the procedures and standards used. The paper has been updated to include this information.**

3. *Page 11,  last paragraph the authors state: "This result strongly suggests that the operator effect in the bullet scanning procedure must be taken into account in order to assume pairwise independence." I don’t understand this statement at all. Do the authors mean independence between features? The pair-wise comparisons between the feature vectors in this study obviously ARE NOT independent (and the features are likely not independent either…). I’m guessing I'm not understanding something; or the author's logic. I suggest the authors expand this paragraph a bit and clarify the statement.*

  **Apologies. The intent here in terms of independence was pairwise independence of scans between the two studies. Because Hamby 252 and Hamby 44 were scans collected on bullets fired from the same gun barrels, assuming that microscope operation and the scanning procedure is precisely defined and perfect, there should be independence between the scans conducted by different operators, and therefore the features derived from these scans should be the same, independent of who conducted the scan. We have clarified this in the manuscript.**

4. *pp14, last paragraph which discusses tank rash on a surface. The authors state: "We extract a signature from the unaffected half and attempt to match this signature to its full known match." Just out of curiosity, how did the authors segment out tank rash? Automatically or subjectively by hand? Could they please state explicitly."*

  **This was determined subjectively by visual inspection, and we have clarified that point in the paper.**

5. *pp18 section 5 “From Lands to Bullets”. A very similar model was already presented by Chu et al in 2010: Chu et al. J Forensic Sci, March 2010, Vol. 55, No. 2.  pp. 341-347. It should be cited and deserves at least a sentence of discussion/comparison...*

  **Thank you. We have added this citation and added some discussion of the differences compared to our approach.**

# Reviewer 2

1. *In this case, I doubt the usability of their method in other real cases or other set of barrels. So saying the features are robust will be only for this specific barrel unless proven in other cases (datasets)*

  **We do agree that much more work needs to be done to assess the performance of the algorithm in real cases, and in the case of other barrels. At the time of writing, further data was not publicly available to us, and that is why we elected to assess what effects were available (operator or study effects in the bullet scanning process), or to simulate conditions that may be more indicative of a real-world scenario (degraded land impressions). The features used were derived and standardized such that we believe they will be applicable to out-of-sample real world data as we continue to obtain it. Indeed, we've seen some promising results with bullet sets sent to us by police departments since the time of writing.**

2. *The partitioning procedure of the data to 80%-20% is unclear. It seems the authors partitioning after getting the all pairwise land-to-land comparisons. This will skew the error rate and won’t represent the true error rate if new barrel/bullet is applied to this algorithm. Instead, try to keep out one bullet from each barrel/ or keep out a whole barrel to assess the error rate correctly.*

  **Thank you for the feedback. We have redone the partitioning scheme and clarified the method in the paper. As you suggested, we have elected to hold out 20% of the barrels, and then used all land-to-land comparisons between lands in these barrels. The results match very closely to the previous scheme, with a very high specificity and a relatively lower sensitivity.**

3. *The overlayed density plots in Figure 3 needs to have better description. Are you looking at all pairwise known non match scores in the figure? What is the point you are trying to make — symmetric and normally distributed vs skewed ? Does the two features differ? in what way? Is one better than the other? In addition, what bandwidth was used. Did you use the default? In that case, it will depend on the number of matches and non matches. I suspect if you use the same bandwidth, the overlap will be different. Please use Empirical - CDF plots instead to avoid the issue of bandwidth selection for the two groups and will give better view of the plots and corresponding error rates.*

  **We have updated these plots to use empirical CDF plots, as per your advice about the bandwidth issue. Thanks for pointing this out. We have also clarified the descriptions of these plots to indicate the relevant information, including the symmetry of each distribution and the differences across the features.**

4. *Try to summarize Figures 12-15 in one figure- side by side boxplots should work fine.*

  **These figures have been updated. Thank you.**

5. *The latter part of the paper is hard to follow. It appears that some of the figures and tables are missing from the paper (See pages 11 and 18). If this is to be published it needs a significant rewrite and reorganization from the current form.*

  **We have updated this section to include the missing figures and tables, and reorganized them as appropriate. We hope you agree the new structure of this section is much easier to follow.**

6. *P 2 paragraph 2 “Real world applications ... “ the statements made here need citations*

  **Citations have been added for these statements.**

7. *P 3 features description: indicate which ones are new features vs already proposed ones*

  **This has now been included.**

8. *P 8 describe data partitioning method more clearly. Make sure the test set is correlated with the validation set.*

  **Thank you, we have clarified the data partitioning used and revised it based on your advice.**

9. *P 8 “It can be seen that Hamby252 to Hamby252 comparisons exhibit the fewest errors, while Hamby252 to Hamby44 comparisons exhibit the most on average” — the ideal would be to have equal error rates in all cases - data partitioning method may solve this issue*

  **The new data partitioning scheme has improved the results, however hasn't completely eliminated this effect. The discussion has been updated to reflect the new results.**

10. *P 9 “... resemble very closely the feature distribution of the true negatives, with a slightly higher average ccf.” add sentence describing what that section is going to be about — then discuss the plot and rest of the subsection*

  **We have added this clarifying sentence.**

11. *P 9 Figure 4 — make better choice of colors and line types for black and white version of the paper. Same with Figure 5*

  **We have updated the figures accordingly, combining them into one figure in the process.**

12. *P 11 missing table*

  **Apologies. This has been corrected.**

12. *P 18 missing figure*

  **This has also been corrected.**

13. *P19 avoid equation number on each line of the equation - this is a straight forward calculation and do not need to be explained line by line - “Where” after the equation should be “where”*

  **Updated. Thank you.**

### 2017-10-03 degraded-paper.Rmd
Paper updates

---
title: "Algorithmic Approaches to Match Degraded Land Impressions"
author: "Eric Hare, Heike Hofmann, Alicia Carriquiry"
date: "9/5/2017"
bibliography: references.bib
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    includes:
      in_header: preamble.tex
fontsize: 12pt
---

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE}
library(ggplot2)
library(xtable)
library(RMySQL)
library(tidyverse)
library(bulletr)
library(gridExtra)
library(randomForest)
library(caret)
library(broom)
library(matrixcalc)

@@

Intense scrutiny has been focused on the process of bullet matching in recent years [e.g., @giannelli:2011]. Bullet matching, the process of determining whether two bullets could have been fired from the same gun barrel, has traditionally been performed without meaningful determination of error rates or statistical assessments of uncertainty [@NAS:2009]. There have been some attempts towards developing mathematical and statistical approaches to bullet matching.One such attempt was the definition of CMS, the Consecutively Matching Striae [@biasotti:1959], with a cutoff of six to separate matches from non-matches. Still, rigorous assessments of the applicability of such cutoffs have not to this point been described [@pcast2016].

Recently, several authors have addressed these well-known shortcomings. Focusing on firing pin impressions and breech faces, @riva:2014 have described an automated algorithm using 3D images that enables comparison between pairs of exemplars. Other examples of work in this and related areas include @petraco:2012, @chu:2011, @vorburger:2011, and others. In our approach to this problem, Automatic Matching of Bullet Lands, we used the Hamby 252 set [@hamby:2009] to train and develop a random forest in order to provide a matching probability for two bullet lands [@2016arXiv160105788H]. While the algorithm had a very strong performance on this set, some limitations were immediately clear. For instance, performance was assessed only on this single set of 35 bullets fired from a consecutively manufactured set of only ten gun barrels. Each of these bullets was part of controlled study, and the full lands were available for matching. While there were some data quality issues, this was still a near ideal test case for the algorithm.

Real world applications of bullet matching often involve the recovery of fragments of bullets from the crime scene. Traditional features used in forensic examination work well for a full land, but there has been less investigation into their performance in the case of a fragmented land. For example, the CMS is naturally limited by the portion of the land that can be recovered.

@@

To start, we introduce a standardized version of each of the features used in the matching routine proposed by [@2016arXiv160105788H]. These features are computed on *aligned pairs of bullet land impressions* rather than on individual lands. This enables us, for instance, to compute the number of matching striae. Table \ref{tab:ccf1} and Table \ref{tab:ccf2} provide an example of six land-to-land comparisons and the derived features for these comparisons. The two `profile_id` columns identify a particular land from the two Hamby datasets, and the remaining columns are the derived features for these comparisons.

```{r, echo=FALSE, results='asis'}
result <- my_ccf %>%
    filter(compare_id == 4) %>%
    head %>%
    collect %>%
    select(2:8) %>%
    as.data.frame

print(xtable(result, digits = c(0, 0, 0, 4, 4, 4, 4, 4), label = "tab:ccf1", caption = "A sample of six land-to-land comparisons, and derived features for these comparisons."), comment = FALSE, include.rownames = FALSE)
```

```{r, echo=FALSE, results='asis'}
result2 <- my_ccf %>%
    filter(compare_id == 4) %>%
    head %>%
    collect %>%
    select(9:ncol(.)) %>%
    as.data.frame

print(xtable(result2, digits = c(4, 4, 4, 4, 4, 4, 4, 4), label = "tab:ccf2", caption = "The remaining derived features for the previous six land-to-land comparisons."), comment = FALSE, include.rownames = FALSE)
```

We generalize the definitions of these features to account for the possibility that we may be handling degraded bullet lands, where only fragments can be recovered. The definition of each feature is given below, where $f(t)$ represents the height values of the first profile, and $g(t)$ the height values of the second:

- **ccf** (%) is the maximum value of the Cross-Correlation function evaluated at the optimal alignment. The Cross-Correlation function is defined as $C(\tau) = \int_{-\infty}^{\infty} f(t)g(t + \tau)dt$ where $\tau$ represents the the lag of the second signature [@vorburger:2011]. 
- **rough_cor** (%) is a new feature that quantifies the correlation between the two signatures after performing a second LOESS smoothing stage and then subtracting the result from the original signatures. This attempts to model the roughness of the surface after removing structure such as waviness.
- **lag** (mm) Is the optimal lag for the ccf value.
- **D** (mm) is the Euclidean vertical distance between each height value of the aligned signatures. This is defined as $D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2$. This is a measure of the total variation between two functions [@clarkson1933definitions].
- **sd_D** (mm) provides the standard deviation of the values of *D* from above.
- **signature_length** (mm) is the overall length of the smallest of the two aligned signatures.
- **overlap** (%) provides the percentage of the two signatures that overlap after the alignment stage.
- **matches** (per mm) is the number of matching peaks/valleys (striae) per millimeter of the overlapping portion of the aligned signatures.
- **mismatches** (per mm) is the number of mismatching peaks/valleys (striae) per millimeter of the overlapping portion of the aligned signatures.
- **cms** (per mm) is the number of consecutively matching peaks/valleys (striae) per millimeter of the overlapping portion of the aligned signatures [@biasotti:1959, @thompson:2013].
- **non_cms** (per mm) is the number of consecutive mismatching peaks/valleys (striae) per millimeter of the overlapping portion of the aligned signatures.
- **sum_peaks** (per mm) is the the sum of the average heights of matched striae.

@@

We can observe the distributions of both CCF and the Roughness Correlation side by side, differentiating between known matches and known non-matches. Figure \@ref(fig:roughcordist) shows this. It can be seen that the separation of the two groups along both variables is quite strong, but many known-matches have low values for each. The distributions of known non-matches, on the other hand, is relatively symmetric and normal.

```{r roughcordist, echo=FALSE, fig.cap='Distributions of the Roughness Correlation compared to the CCF for known matches and known non-matches'}
cor_compare <- CCFs_withlands %>%
  select(ccf, rough_cor, match) %>%
  gather(key = feature, value = value, ccf, rough_cor) %>%
  mutate(match = factor(match, labels = c("Known non-matches (KNM)", "Known matches (KM)")))

qplot(value, group=match, geom="density", data=cor_compare, fill=match, 
      alpha=I(0.6), colour=I("grey20")) + 
  facet_wrap(~feature, nrow = 2) +
  theme_bw() + ylab("") + xlab("") +
  scale_fill_brewer("Bullet Land Pairs", palette="Paired") +
  theme(legend.background = element_rect(colour="grey75"),
        axis.title.y=element_blank(),
        plot.margin = unit(c(0,0,0,0), unit="cm"))
```

@@

Using these features, we can train a randomForest [@randomForest] model which attempts to predict whether two lands match given the value of the features.  There are currently three studies for bullets included in the NIST ballistics database (REF). Those are Hamby (Set 252), Hamby (Set 44), and Cary. For purposes of analysis we describe in this paper, we exclude the Cary bullets from consideration, because the study was designed to assess the persistence of striation markings over a series of fires from the same barrel. Thus, every Cary bullet is a known match to every other Cary bullet, although early firings do not produce the same markings that later firings do. Hence, we will consider Hamby (Set 252) and Hamby (Set 44) only.

We can now train the forest using the features we defined earlier. We split the data into an 80% training, 20% testing sets to assess the algorithm's out of sample performance, using the `caret` package [@caretpkg]. Table \ref{tab:avgforest} displays the results in the form of a confusion matrix on the test set, averaged over ten independent random forests trained on ten random data subsets. There are a total of 83,028 land-to-land comparisons, of which 1208 are among known matching land impressions and 81,820 are among known non-matching land impressions. It can be seen that false positives are exceedingly rare, but false negatives occur more frequently (approximately 65 false negative land to land comparisons on the test set, compared with an average of less than four false positives).

```{r, echo=FALSE, results='asis'}
includes <- setdiff(names(CCFs_withlands), c("compare_id", "profile1_id", "profile2_id",
                                         "study.x", "study.y", "barrel.x", "barrel.y",
                                         "bullet.x", "bullet.y", "land.x", "land.y",
                                         "land_id.x", "land_id.y", "signature_length"))

set.seed(20170222)
trainIndex <- createDataPartition(1:nrow(CCFs_withlands), p = 0.8, list = TRUE, times = 10)
match_result <- trainIndex %>% purrr::map_df(function(index) {
    CCFs_train <- CCFs_withlands[index,]
    CCFs_test = setdiff(CCFs_withlands, CCFs_train)

@@

    mydf <- confmat %>%
        mutate(Study = paste(study.x, study.y, sep = "_"),
               Result = rep(c("True Negative", "False Positive", "False Negative", "True Positive"), times = length(unique(Study)))) %>%
        mutate(Study = replace(Study, which(Study == "Hamby44_Hamby252"), "Hamby252_Hamby44")) %>%
        dplyr::select(Study, Result, Count) %>%
        group_by(Study, Result) %>%
        summarise(Count = sum(Count)) %>%
        as.data.frame

    return(mydf)
}) 
match_result <- cbind(id = rep(1:10, each = nrow(match_result) / 10), match_result)
match_result_wide <- match_result %>%
    spread(key = Result, value = Count)

@@

These results suggest that our algorithm is  too conservative in predicting a match when in fact the bullets were fired from the same gun barrel. We can break down the confusion matrix by the study from which each of the two land impressions originated. Table \ref{tab:avgforeststudy} shows the average confusion matrix for the 10 random forests, broken down by study. It can be seen that Hamby252 to Hamby252 comparisons exhibit the fewest errors, while Hamby252 to Hamby44 comparisons exhibit the most errors on average. This intuitively makes some sense given the potential presence of scanner operator effects, which we address further in this section.

@@

print(xtable(result_wide_row, label = "tab:avgforeststudy", caption = "The average confusion matrix for the 10 random forests, broken down by study. It can be seen that Hamby252 to Hamby252 comparisons exhibit the fewest errors, while Hamby252 to Hamby44 comparisons exhibit the most on average."), comment = FALSE, include.rownames = FALSE, table.placement = "H")
```

@@

As a first stage to assessing feature robustness, we produce parallel coordinate plots of the various features based on true positives, true negatives, false positives, and false negatives. Figures \ref{fig:pcp} and \ref{fig:pcp2} display these plots. The means of the true positive and the false negative groups are shown, respectively, in the two figures. False positives tend to occur with anomalously high values of `sum_peaks`, `matches`, and `ccf`, while false negatives tend to resemble very closely the feature distribution of the true negatives, with a slightly higher average `ccf`.

```{r pcp, echo=FALSE, message=FALSE, fig.cap="Parallel coordinate plot of the features based on the random forest confusion matrix for true and false positives.", fig.height=6, fig.width=10}
set.seed(20170222)
CCFs_train <- sample_frac(CCFs_withlands, size = .8)
CCFs_test = setdiff(CCFs_withlands, CCFs_train)

includes <- setdiff(names(CCFs_withlands), c("compare_id", "profile1_id", "profile2_id",
                                         "study.x", "study.y", "barrel.x", "barrel.y",
                                         "bullet.x", "bullet.y", "land.x", "land.y",
                                         "land_id.x", "land_id.y", "signature_length", "lag", "overlap"))

@@

ggplot() +
  geom_line(data = filter(CCFs_tall, Type %in% c("True Positive", "False Positive")), aes(x = Feature, y = Value, colour = Type, group = id, alpha = Type, size = Type)) +
  geom_line(data = filter(pcp_data_tall, Type %in% c("True Positive")), aes(x = Feature, y = Value, colour = Type, group = Type), size = 3) +
  ylab("Normalized Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c("#e60000", "#000099")) +
  scale_alpha_manual(values = c(1, .05)) +
  scale_size_manual(values = c(1, 1))
```

```{r pcp2, echo=FALSE, message=FALSE, fig.cap="Parallel coordinate plot of the features based on the random forest confusion matrix for true and false negatives.", fig.height=6, fig.width=10}
mysamp <- sample(CCFs_tall$id[CCFs_tall$Type == "True Negative"], 1000)
trueneg_sample <- CCFs_tall %>%
  filter(id %in% mysamp) %>%
  rbind(CCFs_tall %>% filter(Type == "False Negative"))

ggplot() +
  #geom_line(data = filter(CCFs_tall, Type %in% c("True Negative", "False Negative")), aes(x = Feature, y = Value, colour = Type, group = id, alpha = Type, size = Type)) +
  geom_line(data = trueneg_sample, aes(x = Feature, y = Value, colour = Type, group = id, alpha = Type)) +
  theme_bw() +
  geom_line(data = filter(pcp_data_tall, Type %in% c("False Negative")), aes(x = Feature, y = Value, colour = Type, group = Type), size = 3) +
  ylab("Normalized Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c("#ff9999", "#b3b3ff")) +
  scale_alpha_manual(values = c(.5, .1))
```

@@

We attempt to quantify the effect of the study on the matching probability by fitting a new random forest which is designed to predict the study based on the derived features. Ideally, if the assumption of independence between lands holds across different operators, this forest should have poor performance - The set of derived features should be relatively consistent among known matches and known non-matches regardless of the study since the Hamby data in both sets originated from the same gun barrels.

```{r, echo=FALSE}
includes_study <- setdiff(names(CCFs_withlands), c("compare_id", "profile1_id", "profile2_id",
                                         "barrel.x", "barrel.y", "match",
                                         "bullet.x", "bullet.y", "land.x", "land.y",
                                         "land_id.x", "land_id.y", "study.x", "study.y", "signature_length"))

result_study <- trainIndex %>% purrr::map_df(function(index) {
    CCFs_train <- CCFs_withlands[index,]
    CCFs_test = setdiff(CCFs_withlands, CCFs_train)

@@

    return(confmat)
})
```

>>>>

---
title: "Algorithmic Approaches to Match Degraded Land Impressions"
author: "Eric Hare, Heike Hofmann, Alicia Carriquiry"
date: "10/05/2017"
bibliography: references.bib
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    includes:
      in_header: preamble.tex
fontsize: 12pt
---

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE}
library(ggplot2)
library(xtable)
library(RMySQL)
library(tidyverse)
library(bulletr)
library(gridExtra)
library(randomForest)
library(caret)
library(broom)
library(matrixcalc)
library(parallel)

@@

Intense scrutiny has been focused on the process of bullet matching in recent years [e.g., @giannelli:2011]. Bullet matching, the process of determining whether two bullets could have been fired from the same gun barrel, has traditionally been performed without meaningful determination of error rates or statistical assessments of uncertainty [@NAS:2009]. There have been some attempts towards developing mathematical and statistical approaches to bullet matching. One such attempt was the definition of CMS, the Consecutively Matching Striae [@biasotti:1959], with a cutoff of six to separate matches from non-matches. Still, rigorous assessments of the applicability of such cutoffs have not to this point been described [@pcast2016].

Recently, several authors have addressed these well-known shortcomings. Focusing on firing pin impressions and breech faces, @riva:2014 have described an automated algorithm using 3D images that enables comparison between pairs of exemplars. Other examples of work in this and related areas include @petraco:2012, @chu:2011, @vorburger:2011, and others. In our approach to this problem, Automatic Matching of Bullet Lands, we used the Hamby 252 set [@hamby:2009] to train and develop a random forest in order to provide a matching probability for two bullet lands [@2016arXiv160105788H]. While the algorithm had a very strong performance on this set, some limitations were immediately clear. For instance, performance was assessed only on this single set of 35 bullets fired from a consecutively manufactured set of only ten known and 15 unknown gun barrels. Each of these bullets was part of controlled study, and the full lands were available for matching. While there were some data quality issues, this was still a near ideal test case for the algorithm.

Real world applications of bullet matching often involve the recovery of fragments of bullets from the crime scene [@national2004forensic]. Traditional features used in forensic examination work well for a full land, but there has been less investigation into their performance in the case of a fragmented land. For example, the CMS is naturally limited by the portion of the land that can be recovered and varies across manufacturers [@chu:2011].

@@

To start, we introduce a standardized version of each of the features used in the matching routine proposed by [@2016arXiv160105788H]. These features are computed on *aligned pairs of bullet land impressions* rather than on individual lands. This enables us, for instance, to compute the number of matching striae between two lands. We generalize the definitions of these features to account for the possibility that we may be handling degraded bullet lands, where only fragments can be recovered. The definition of each feature is given below, where $f(t)$ represents the height values of the first profile at position $t$ along the signature, and $g(t)$ the height values of the second. An indication of whether the feature is new since @2016arXiv160105788H is also given:

- **ccf** (%) is the maximum value of the Cross-Correlation function evaluated at the optimal alignment. The Cross-Correlation function is defined as $C(\tau) = \int_{-\infty}^{\infty} f(t)g(t + \tau)dt$ where $\tau$ represents the the lag of the second signature [@vorburger:2011]. 
- **rough_cor** (new) (%) feature that quantifies the correlation between the two signatures after performing a second LOESS smoothing stage and then subtracting the result from the original signatures. This attempts to model the roughness of the surface after removing structure such as waviness.
- **lag** (mm) Is the optimal lag for the ccf value.
- **D** (mm) is the Euclidean vertical distance between each height value of the aligned signatures. This is defined as $D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2$. This is a measure of the total variation between two functions [@clarkson1933definitions].
- **sd_D** (mm) provides the standard deviation of the values of *D* from above.
- **signature_length** (mm) is the overall length of the smallest of the two aligned signatures.
- **overlap** (new) (%) provides the percentage of the two signatures that overlap after the alignment stage.
- **matches** (per mm) is the number of matching peaks/valleys (striae) per millimeter of the overlapping portion of the aligned signatures.
- **mismatches** (per mm) is the number of mismatching peaks/valleys (striae) per millimeter of the overlapping portion of the aligned signatures.
- **cms** (per mm) is the number of consecutively matching peaks/valleys (striae) per millimeter of the overlapping portion of the aligned signatures [@biasotti:1959, @thompson:2013].
- **non_cms** (per mm) is the number of consecutive mismatching peaks/valleys (striae) per millimeter of the overlapping portion of the aligned signatures.
- **sum_peaks** (per mm) is the the sum of the average heights of matched striae.

@@

We can observe the distributions of both CCF and the Roughness Correlation side by side, differentiating between known matches and known non-matches. Figure \@ref(fig:roughcordist) displays this as an empirical CDF plot. It can be seen that the separation of known matches and known non-matches along both the CCF and the Roughness Correlation is quite strong and follows similar distributions (the known non-matches are relatively symmetric, while the known matches are very skewed left). However, some known non-matches with CCF values that would typically be indicative of a match have relatively lower values for the Roughness Correlation, which indicates that this feature could provide some added value when it comes to discriminating between matches and non-matches.

```{r roughcordist, echo=FALSE, fig.cap='Empirical CDFs of the Roughness Correlation compared to the CCF for known matches and known non-matches. It can be seen that the distribution of each feature for the known non-matches is quite symmetric, while the distribution for each feature for the known matches is skewed left.'}
cor_compare <- CCFs_withlands %>%
  select(ccf, rough_cor, match) %>%
  gather(key = feature, value = value, ccf, rough_cor) %>%
  mutate(match = factor(match, labels = c("Known non-matches (KNM)", "Known matches (KM)")))

ggplot(data = cor_compare, aes(x = value, color = match)) +
  stat_ecdf() +
  facet_wrap(~feature, nrow = 2) +
  theme_bw() + ylab("") + xlab("") +
  scale_color_brewer("Bullet Land Pairs", palette="Paired") +
  theme(legend.background = element_rect(colour="grey75"),
        axis.title.y=element_blank(),
        plot.margin = unit(c(0,0,0,0), unit="cm"))
```

@@

Using these features, we can train a randomForest [@randomForest] model which attempts to predict whether two lands match given the value of the features. There are currently three studies for bullets included in the NIST Ballistics Toolmark Research Database. Those are Hamby (Set 252), Hamby (Set 44), and Cary. For purposes of the analysis we describe in this paper, we exclude the Cary bullets from consideration, because the study was designed to assess the persistence of striation markings over a series of fires from the same barrel. Thus, every Cary bullet is a known match to every other Cary bullet. Hence, we will consider Hamby (Set 252) and Hamby (Set 44) only. This leaves us a total of 83,028 land-to-land comparisons, of which 1208 are among known matching land impressions and 81,820 are among known non-matching land impressions.

We can now train the forest using the features we defined earlier. Using the `caret` package [@caretpkg], we perform the following partitioning scheme. Out of the 50 barrels total, ten knowns and fifteen unknowns from each of the two Hamby sets, we hold out ten barrels randomly as a testing set, and use the remaining 40 to train the model. We repeat this procedure ten times and average the confusion matrix in order to assess the model accuracy with different holdout samples. Table \ref{tab:avgforest} displays the results in the form of a confusion matrix on the test set, averaged over these ten independent random forests trained on ten random barrel subsets. It can be seen that false positives are exceedingly rare, but false negatives occur more frequently (approximately 21` false negative land to land comparisons on the test set, compared with an average of less than two false positives).

```{r, echo=FALSE, results='asis'}
includes <- setdiff(names(CCFs_withlands), c("compare_id", "profile1_id", "profile2_id",
                                         "study.x", "study.y", "barrel.x", "barrel.y",
                                         "bullet.x", "bullet.y", "land.x", "land.y",
                                         "land_id.x", "land_id.y", "signature_length"))
  CCFs_withlands <- CCFs_withlands %>%
  mutate(barrel_id.x = interaction(study.x, barrel.x),
         barrel_id.y = interaction(study.y, barrel.y))

set.seed(20170222)
mysamps <- replicate(10, expr = {sample(unique(CCFs_withlands$barrel_id.x), ceiling( length(unique(CCFs_withlands$barrel_id.x)) / 5))}, simplify = FALSE)

cl <- makeCluster(10, renice = 0)
clusterExport(cl, varlist = c("CCFs_withlands", "includes"))
match_result <- parLapply(cl, mysamps, function(mysamp) {
  library(dplyr)
  library(randomForest)

  CCFs_train <- CCFs_withlands %>%
    filter(!(barrel_id.x %in% mysamp),
         !(barrel_id.y %in% mysamp))

  CCFs_test <- CCFs_withlands %>%
    filter(barrel_id.x %in% mysamp,
         barrel_id.y %in% mysamp)

@@

    mydf <- confmat %>%
        mutate(Study = paste(study.x, study.y, sep = "_"),
               Result = rep(c("True Negative", "False Positive", "False Negative", "True Positive"), times = length(unique(Study)))) %>%
        mutate(Study = replace(Study, which(Study == "Hamby44_Hamby252"), "Hamby252_Hamby44")) %>%
        dplyr::select(Study, Result, Count) %>%
        group_by(Study, Result) %>%
        summarise(Count = sum(Count)) %>%
        as.data.frame
})

match_result <- do.call(rbind, match_result)

match_result <- cbind(id = rep(1:10, each = nrow(match_result) / 10), match_result)
match_result_wide <- match_result %>%
    spread(key = Result, value = Count)

@@

These results suggest that our algorithm is too conservative in predicting a match when in fact the bullets were fired from the same gun barrel. We can break down the confusion matrix by the study from which each of the two land impressions originated. Table \ref{tab:avgforeststudy} shows the average confusion matrix for the 10 random forests, broken down by study. It can be seen that Hamby252 to Hamby252 comparisons exhibit the fewest errors, while Hamby44 to Hamby44 comparisons exhibit the most errors on average. This intuitively makes some sense given the potential presence of scanner operator effects, which we address further in this section.

@@

print(xtable(result_wide_row, label = "tab:avgforeststudy", caption = "The average confusion matrix for the 10 random forests, broken down by study. It can be seen that Hamby252 to Hamby252 comparisons exhibit the fewest errors, while Hamby44 to Hamby44 comparisons exhibit the most on average."), comment = FALSE, include.rownames = FALSE, table.placement = "H")
```

@@

Our goal is to assess the robustness of the previously defined features as it pertains to our bullet matching routines. This goal is both a backward looking assessment of our previous results for full land-to-land comparisons, and a forward looking one to help support the case that these can be used in the degraded land case. As a first stage to assessing this robustness, we produce parallel coordinate plots of the various features based on true positive, true negative, false positive, and false negative land-to-land matches. Figure \ref{fig:pcp} displays these plots. The means of the true positive and the true negative groups are shown as thick blue lines, respectively, in the two panels. The dashed lines represent individual land to land comparisons, with errors highlighted larger in red. It can be seen that the few false positives tendt o have anomalously high `ccf` or `matches`, while the false negatives tend have a lot of variability, though tending to also have a high `ccf` value.

```{r pcp, echo=FALSE, message=FALSE, fig.cap="Parallel coordinate plot of the features based on the random forest confusion matrix for true and false positives (above), and true and false negatives (below). False positives tend to have some feature anomalously high, while false negatives exhibit quite a spread, sometimes having very large values of CCF or CMS, for example.", fig.height=6, fig.width=10}
set.seed(20170222)
CCFs_train <- sample_frac(CCFs_withlands, size = .8)
CCFs_test = setdiff(CCFs_withlands, CCFs_train)

includes <- setdiff(names(CCFs_withlands), c("compare_id", "profile1_id", "profile2_id",
                                         "study.x", "study.y", "barrel.x", "barrel.y",
                                         "bullet.x", "bullet.y", "land.x", "land.y",
                                         "land_id.x", "land_id.y", "signature_length", "lag", "overlap", "barrel_id.x", "barrel_id.y"))

@@

p1 <- ggplot() +
  geom_line(data = filter(CCFs_tall, Type %in% c("True Positive")), aes(x = Feature, y = Value, colour = Type, group = id), alpha = 0.1, size = 0.2, linetype = "dashed") +
  geom_line(data = filter(CCFs_tall, Type %in% c("False Positive")), aes(x = Feature, y = Value, colour = Type, group = id), alpha = 1, size = 0.8, linetype = "dashed") +
  geom_line(data = filter(pcp_data_tall, Type %in% c("True Positive")), aes(x = Feature, y = Value, colour = Type, group = Type), size = 2) +
  ylab("Normalized Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c("#e60000", "#000099")) +
  scale_size_manual(values = c(1, 1))

set.seed(20170222)
mysamp1 <- sample(CCFs_tall$id[CCFs_tall$Type == "True Negative"], 226)
mysamp2 <- sample(CCFs_tall$id[CCFs_tall$Type == "False Negative"], 20)
trueneg_sample <- CCFs_tall %>%
  filter(id %in% mysamp1)
falseneg_sample <- CCFs_tall %>% 
  filter(id %in% mysamp2)

p2 <- ggplot() +
  #geom_line(data = filter(CCFs_tall, Type %in% c("True Negative", "False Negative")), aes(x = Feature, y = Value, colour = Type, group = id, alpha = Type, size = Type)) +
  geom_line(data = trueneg_sample, aes(x = Feature, y = Value, colour = Type, group = id), alpha = 0.1, size = 0.2, linetype = "dashed") +
  geom_line(data = falseneg_sample, aes(x = Feature, y = Value, colour = Type, group = id), alpha = 1, size = 0.4, linetype = "dashed") +
  theme_bw() +
  geom_line(data = filter(pcp_data_tall, Type %in% c("True Negative")), aes(x = Feature, y = Value, colour = Type, group = Type), size = 2) +
  ylab("Normalized Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = c("#e60000", "#000099"))

grid.arrange(p1, p2, nrow = 2)
```

@@

We attempt to quantify the effect of the study on the matching probability by fitting a new random forest which is designed to predict the study from which the scans came from based on the derived features. It should be noted that the two sets of scans were performed by two trained professionals at NIST, and therefore this analysis is intended to shed light on how slightly different operating procedures for the scans may lead to varying results in algorithms derived from these scans. Ideally, if the assumption of independence between lands holds across different operators, this forest should have poor performance - The set of derived features should be relatively consistent among known matches and known non-matches regardless of the study since the Hamby data in both sets originated from the same gun barrels.

```{r, echo=FALSE}
includes_study <- setdiff(names(CCFs_withlands), c("compare_id", "profile1_id", "profile2_id",
                                         "barrel.x", "barrel.y", "match",
                                         "bullet.x", "bullet.y", "land.x", "land.y",
                                         "land_id.x", "land_id.y", "study.x", "study.y", "barrel_id.x", "barrel_id.y", "signature_length"))

clusterExport(cl, varlist = c("CCFs_withlands", "includes_study"))

result_study <- parLapply(cl, mysamps, function(mysamp) {
  library(dplyr)
  library(randomForest)

  CCFs_train <- CCFs_withlands %>%
    filter(!(barrel_id.x %in% mysamp),
         !(barrel_id.y %in% mysamp))

  CCFs_test <- CCFs_withlands %>%
    filter(barrel_id.x %in% mysamp,
         barrel_id.y %in% mysamp)

@@

    return(confmat)
})

stopCluster(cl)
```
